Higher order unification for free!

### 2024-04-09 paper/main.tex
rename properties

\begin{proposition}[Properties of \hrun outside \llambda]
\begin{gather}
  \exists \rho, \rho s_1 \Eo \rho s_2 \Rightarrow
  \C{s_i}{t_i}{m_i}{l_i} \Rightarrow
  t_1 \Ue t_2 \mapsto \sigma \label{prop:compilation-nllam}\\
  \sigma_{p-1} \hoUnifPb_{p} \in \llambda \Rightarrow \label{prop:step-nllam}\\
  \exists \rho,\stepF{\foUnifPb}{p}{\rho}{\rho'} \Leftrightarrow
  \stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_{p}}{\linkStore_p} \nonumber
\end{gather}
\end{proposition}

@@

\begin{proposition}[\wellb{}-preservation of \hstep]\label{prop:nf}
$\forall \hoUnifPb, \forall \linkStore, \forall p, \forall \sigma, \forall \sigma'$
$$
\begin{array}{l}
\wellb(\sigma\hoUnifPb) \land
  \sigma\hoUnifPb_{p_l} \Ue \sigma\hoUnifPb_{p_r} \mapsto {\sigma'}
  \Rightarrow \wellb(\sigma' \hoUnifPb)\\
\wellb(\sigma\hoUnifPb) \land
  \progress(\linkStore,\sigma) \mapsto (\_,\sigma')
  \Rightarrow \wellb(\sigma' \hoUnifPb)
\end{array}
$$
\end{proposition}

\noindent
A less formal way to state \ref{prop:nf} is that \hstep{} and \progress never
``commit'' an unneeded $\lambda$-abstraction in $\sigma$ (a $\lambda$
that could be erased by an $\eta$-contraction),
nor put in $\sigma$ an application whose head is flexible
(an application node that could be erased by a $\beta$-reduction).

@@

\newcommand{\progressetadedup}{\emph{progress-$\eta$-deduplicate}}
\begin{definition}[\progressetadedup]\label{def:progressetadedup}
  A link \linketaM{\Gamma}{X_{\vec{s}}}{T} is removed from \linkStore when
  another link \linketaM{\Delta}{X_{\vec{r}}}{T'} is in  \linkStore.
  By \cref{inv:uvaarity} the length of $\vec{s}$ and $\vec{r}$ is the same
  hence we can move the term $T'$ from $\Delta$ to $\Gamma$ by renaming its
  bound variables, i.e. $T'' = T'[\vec{r}/\vec{s}]$.
  We then run $T \Ue{} T''$ (under the context $\Gamma$).
\end{definition}

\noteD[inline]{Below the proof of \cref{prop:nf}, ho usato 3 lemmi ausiliari,
forse si può compattare in una prova più piccola?}
\begin{lemma}
  Given a \linketa $l$, the unification done by \progressetaleft is between
  terms in $\wellb$  
  \label{lemma:unif-eta-1}
\end{lemma}

\begin{proof}[Proof sketch]
  Let $\sigma$ be the substitution, such that $\wellb(\sigma)$. $\lhs \in
  \sigma$, therefore $\wellb(\lhs)$. By \cref{def:progressetaleft}, if 1) \lhs is
  a name, a constant of an application, then, \lhs is unified with the
  $\eta$-reduced term $t$ obtain from \rhs. By \cref{cor:rhs-eta-onelamb}, \rhs
  has one lambda, therefore $\wellb(t)$. Otherwise, 2) \lhs has \elpiIn{lam} as
  functor, \rhs should not be an $\eta$-expansion ans, so, $\wellb(\rhs)$. In
  both cases, unification is performed between terms in \wellb.
\end{proof}

@@

\begin{proof}
  Trivial, since the unification is done between unification variables, which
  are by definition in \wellb.
\end{proof}

\begin{lemma}
  \Cref{prop:nf} holds, i.e., given a substitution $\sigma$ and a \linketa $l$,
  after the activation of $l$, $\wellb(\sigma)$ holds.
\end{lemma}

\begin{proof}[Proof sketch]
  By \cref{lemma:unif-eta-1,lemma:unif-eta-2,lemma:unif-eta-3}, every
  unification performed by the activation of a \linketa is performed between
  terms in \wellb, therefore, the substitution remains \wellb.
  \noteD[inline]{Bisogna aggiungere un lemma nella \cref{sec:nutshell} che dice che
  unificare due termini in \wellb, in una $\sigma$, tale che $\wellb(\sigma)$, 
  non invalida \wellb}
\end{proof}

>>>>

\begin{proposition}[Fidelity recovery]
\begin{gather}
  \exists \rho, \rho s_1 \Eo \rho s_2 \Rightarrow
  \C{s_i}{t_i}{m_i}{l_i} \Rightarrow
  t_1 \Ue t_2 \mapsto \sigma \label{prop:compilation-nllam}\\
  \sigma_{p-1} \hoUnifPb_{p} \in \llambda \Rightarrow \label{prop:step-nllam}\\
  \exists \rho,\stepF{\foUnifPb}{p}{\rho}{\rho'} \Leftrightarrow
  \stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_{p}}{\linkStore_p} \nonumber
\end{gather}
\end{proposition}

@@

\begin{proposition}[\wellb{}-preservation]\label{prop:nf}
$\forall \hoUnifPb, \forall \linkStore, \forall p, \forall \sigma, \forall \sigma'$
$$
\begin{array}{l}
\wellb(\sigma\hoUnifPb) \land
  \sigma\hoUnifPb_{p_l} \Ue \sigma\hoUnifPb_{p_r} \mapsto {\sigma'}
  \Rightarrow \wellb(\sigma' \hoUnifPb)\\
\wellb(\sigma\hoUnifPb) \land
  \progress(\linkStore,\sigma) \mapsto (\_,\sigma')
  \Rightarrow \wellb(\sigma' \hoUnifPb)
\end{array}
$$
\end{proposition}

\noindent
A less formal way to state \ref{prop:nf} is that \hstep{} and \progress never
``commit'' an unneeded $\lambda$-abstraction in $\sigma$ (a $\lambda$
that could be erased by an $\eta$-contraction),
nor put in $\sigma$ an application whose head is flexible
(an application node that could be erased by a $\beta$-reduction).
\todo{dire che W preservation is key for our claims}

@@

\newcommand{\progressetadedup}{\emph{progress-$\eta$-deduplicate}\xspace}
\begin{definition}[\progressetadedup]\label{def:progressetadedup}
  A link \linketaM{\Gamma}{X_{\vec{s}}}{T} is removed from \linkStore when
  another link \linketaM{\Delta}{X_{\vec{r}}}{T'} is in  \linkStore.
  By \cref{inv:uvaarity} the length of $\vec{s}$ and $\vec{r}$ is the same
  hence we can move the term $T'$ from $\Delta$ to $\Gamma$ by renaming its
  bound variables, i.e. $T'' = T'[\vec{r}/\vec{s}]$.
  We then run $T \Ue{} T''$ (under the context $\Gamma$).
\end{definition}

% \noteD[inline]{Below the proof of \cref{prop:nf}, ho usato 3 lemmi ausiliari,
% forse si può compattare in una prova più piccola?}
\begin{lemma}
  Given a \linketa $l$, the unification done by \progressetaleft is between
  terms in $\wellb$  
  \label{lemma:unif-eta-1}
\end{lemma}

\begin{proof}[Proof sketch]
  Let $\sigma$ be the substitution, which is  $\wellb(\sigma)$ (by
  \cref{prop:nf}). $\lhs \in \sigma$, therefore $\wellb(\lhs)$. By
  \cref{def:progressetaleft}, if 1) \lhs is a name, a constant of an
  application, then, $\lambda x.\lhs \appsep x$ is unified with \rhs. By
  \cref{inv:link-eta-right}, $\rhs = \lambda x. t$, therefore $\wellb(t)$.
  Otherwise, 2) \lhs has \elpiIn{lam} as functor, \rhs should not be an
  $\eta$-expansion, so, $\wellb(\rhs)$. In both cases, unification is
  performed between terms in \wellb.
\end{proof}

@@

\begin{proof}
  The unification is done between the \rhs of two
  \linketa. Both \rhs has the shape $\lambda x.t$, with $\wellb(t)$, therefore,
  the unification is done between well-behaved terms.
\end{proof}

\begin{lemma}
  The introduction of \linketa guarantees \fullref{prop:nf}
  \label{lemma:unif-wellb}
\end{lemma}

\begin{proof}[Proof sketch]
  By \cref{lemma:unif-eta-1,lemma:unif-eta-2,lemma:unif-eta-3}, every
  unification performed by the activation of a \linketa is done between
  terms in \wellb, therefore, the substitution remains \wellb.
  \noteD[inline]{Bisogna aggiungere un lemma nella \cref{sec:nutshell} che dice che
  unificare due termini in \wellb, in una $\sigma$, tale che $\wellb(\sigma)$, 
  non invalida \wellb}
\end{proof}
